Type-Safe Modular Parsing

### 2017-01-12 paper/sections/RelatedWork.tex
Some revision on Related Work

Our work integrates several components including extensible parsing,
parser combinators and extensibility techniques. There has been a
great amount of related papers on those hot topics. Some
inspired us of this paper and encourage us for more exploration. This
section will try to lead a discussion on what difference we have made.

\paragraph{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, among them parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with their affects on the performance.
\bruno{what performance issue?}\haoyuan{Revised}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which did not focus on separate
compilation.\bruno{focused little or did not have it at all?}.\haoyuan{The authors have
mentioned separate compilation as a possible yet unimportant feature; they decided to discard it.}
Racket adopts macro system for library-based language extensibility. It is powerful with attributed
ASTs for contextual information, and extensions can be integrated in a modular way. Whereas
such modularity is not flexible enough for language unification, as the syntax is only built from extensions.
Moreover, existing macros cannot be further changed after definition. \haoyuan{correct?}
On the other hand, extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition.
%\bruno{Have you read
%Languages as Libraries? That is probably an important reference, which
%I think we should cite.}
%\haoyuan{what about
%  metafront? it is a macro system but does it have type safety?}
%\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
%  syntax" proposes a system for extensible syntax, where users write
%  EDSLs in their language with concrete syntax. Users can write rules
%  for type-based disambiguation. But separate compilation is again not
%  mentioned. Shall we mention that thesis?} \haoyuan{attribute
%  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, where grammars are compiled
to the generation of parse tables. Those parse tables are expressed as DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive in terms of performance.
Furthermore, our approach supports both \emph{separate compilation}
as well as \emph{modular type-checking}. The extensibility of
parsing is further available at language composition
and lexical level.
%\haoyuan{we do not have explicit
%  correspondence/relationship between abstract syntax and the parser.}

\paragraph{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination, back-tracking and
longest match composition issues, as we discussed before.

Some more recent papers~\cite{Ford2002,Might2011,Frost2008} proposed a series of
novel parsing techniques that addressed the issue
of left-recursion. We have selected Packrat
Parsing for our prototype implementation, as the combinators are
convenient to use in Scala, and Object Algebras can be naturally realized with
OO inheritance. Also, its support
for direct left-recursion is already helpful for designing real-world
parsers practically, though~\cite{warth2008} further demonstrated that
general left-recursion is supported theoretically.
Yet we have mentioned that the components used in our approach
can have alternatives to achieve better modular parsing.

>>>>

Our work touches upon several topics including extensible parsing,
parser combinators and extensibility techniques. However, as far as we
know there's no work that discusses how to do statically type-safe and 
separately compilable modular parsing.

\begin{comment}
There has been a
great amount of related papers on those topics. Some
inspired us of this paper and encourage us for more exploration. This
section will try to lead a discussion on what difference we have made.
\end{comment}

\paragraph{Syntactically Extensible Parsing} Extensible parsing is
used in many different areas. Extensible parser generators are a
mainstream area specially designed for modular syntax and
parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new modules where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars.  NOA~\cite{Gouseti2014} uses Java
annotation processing to gather information about grammars in Java
code. After all information is collected NOA merges all grammar
components, produce an ANTLR~\cite{} grammar, and generates
corresponding parsing code from the ANTLR grammar.  Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In constrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking.

Macro systems like the C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar for
Java using library imports. Composition of syntactic sugar is easy for
users, but it requires many rounds of parsing and adaption. This
significantly affects efficiency of compilation. Since the
implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, the approach does not support separate
compilation. Racket adopts a macro system for library-based language
extensibility~\cite{Tobin-Hochstadt2011}. This is quite powerful and it uses 
attributed ASTs for contextual
information. Moreover extensions can be integrated in a modular
way. However such modularity is not flexible enough for language
unification, as the syntax is only built from extensions.  Moreover,
existing macros cannot be further changed after
definition. \haoyuan{correct?}  Extensible
compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. They focus on the
extensions to a host language. Our library is designed for type-safe 
modular language parsers (not just language 
extensions), with very flexible parser composition.

%\bruno{Have you read
%Languages as Libraries? That is probably an important reference, which
%I think we should cite.}
%\haoyuan{what about
%  metafront? it is a macro system but does it have type safety?}
%\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
%  syntax" proposes a system for extensible syntax, where users write
%  EDSLs in their language with concrete syntax. Users can write rules
%  for type-based disambiguation. But separate compilation is again not
%  mentioned. Shall we mention that thesis?} \haoyuan{attribute
%  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that relates to
separate compilation. 
\textit{Parse table composition}~\cite{Bravenboer2009} 
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach 
is quite different from ours, since it uses parse
 tables, wheres we use parser combinators. 
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}. Moreover, the extensibility of parsing is further
available at language composition and lexical level.

%\haoyuan{we do not have explicit
%  correspondence/relationship between abstract syntax and the parser.}

\paragraph{Parser Combinators} Since Burge~\cite{burge1975} firstly
introduced parser generators, and after Wadler popularized them 
in functional programming~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is 
perhaps the most popular parser combinator library in this line. 
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination, 
high cost for backtracking and longest match composition issues, 
as we discussed in Section~\ref{}. Those limitation make Parsec 
inadequate for modular parsing.

Some more recent work on parser
combinators~\cite{Ford2002,Might2011,Frost2008} proposed a series of
novel parsing techniques that address the issue of
left-recursion. We have selected Packrat parsing for our approach, 
as the combinators are convenient to use in Scala, and 
integrate well with multiple trait inheritance and
Object Algebras. 
%Also,
%its support for direct left-recursion is already helpful for designing
%real-world parsers practically, though~\cite{warth2008} further
%demonstrated that general left-recursion is supported theoretically.
Despite our choice of Packrat parsing, as argued in Section~\ref{}, 
any parser combinator approach which supports left-recursion, longest 
match composition, and makes the cost of backtracking reasonable,
should be applicable to modular parsing.
%Yet we have mentioned that the components used in our approach can
%have alternatives to achieve better modular parsing.
